---
title: "Week 5: ACS Census Data (2015)"
date: 2018-04-30T09:45:50-07:00
draft: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(maps)
require(dplyr)
require(plotly)
require(spotifyr)
```

# Setup 
```{r,eval=F}
require(ggplot2)
require(maps)
require(dplyr)
require(plotly)
require(spotifyr)
```

# Week 5 - County-level American Community Survey (5-year estimates) 2015
[RAW DATA](https://github.com/rfordatascience/tidytuesday/blob/master/data/acs2015_county_data.csv)  
[DataSource: census.gov](https://factfinder.census.gov/faces/nav/jsf/pages/index.xhtml)  
[Kaggle source](https://www.kaggle.com/muonneutrino/us-census-demographic-data)  

This week I am taking inspiration from the Tidy Tuesday submissions of @AidoBo and @jakekaupp.

I'm slightly tweaking @AidoBo's function to plot continuous variables on a map to help me explore.

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">For <a href="https://twitter.com/hashtag/TidyTuesday?src=hash&amp;ref_src=twsrc%5Etfw">#TidyTuesday</a> I created simple function which allows you to plot any continuous variable in the data on a map <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://twitter.com/hashtag/r4ds?src=hash&amp;ref_src=twsrc%5Etfw">#r4ds</a> <a href="https://t.co/6Q1I121VqI">pic.twitter.com/6Q1I121VqI</a></p>&mdash; Aidan Boland (@AidoBo) <a href="https://twitter.com/AidoBo/status/991338257391804416?ref_src=twsrc%5Etfw">May 1, 2018</a></blockquote>
  <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
  
And inspired by @jakekaupp's work showing commute time in terms of number of Despacito listens

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">A blog post catching up on week 4 and week 5 of <a href="https://twitter.com/hashtag/TidyTuesday?src=hash&amp;ref_src=twsrc%5Etfw">#TidyTuesday</a> <a href="https://t.co/AoXuNI5s0j">https://t.co/AoXuNI5s0j</a>  Code available at <a href="https://t.co/kuJdBQG4pn">https://t.co/kuJdBQG4pn</a> <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://twitter.com/hashtag/r4ds?src=hash&amp;ref_src=twsrc%5Etfw">#r4ds</a> <a href="https://t.co/IXjONQ0LXs">pic.twitter.com/IXjONQ0LXs</a></p>&mdash; Jake Kaupp (@jakekaupp) <a href="https://twitter.com/jakekaupp/status/992128540396138496?ref_src=twsrc%5Etfw">May 3, 2018</a></blockquote>
  <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
 

I wanted to adapt the function from above (thanks @AidoBo) to make a commuting map for any song. We can use the spotifyr package to access the length of a given song.
  
  
```{r}
counties= map_data("county")
state=map_data("state")

  county_plot <-function(x){
  ## adapted from
  
  ##https://twitter.com/AidoBo/status/991338257391804416
  
  all_county$x<-all_county[,x] ## a different fix for this? something like aes_string?
  
  ggplot(data=counties,mapping=aes(x=long,y=lat,group=group))+
    geom_polygon(data=all_county, aes(fill=x),color="grey")+labs(fill=x)+scale_fill_distiller(palette="Spectral")+theme_void()+
    geom_path(data=state, aes(x=long,y=lat,group=group),color="black") ## add state boundaries
  
}
  
```

```{r}
setwd("~/Desktop/tidytuesday/data/2018-04-30")
acs<-read.csv("week5_acs2015_county_data.csv")

head(acs)
names(acs)

## from @AidoBo too
all_county<-inner_join(counties,acs %>% mutate(County=tolower(County),State=tolower(State)),by=c("subregion"="County","region"="State"))

```

Get your own Client ID and Client Secret [here](https://beta.developer.spotify.com/documentation/web-api/).

```{r,eval=F}
spotify_client_id="" ## put yours here
spotify_client_secret="" ## put yours here
access_token <- get_spotify_access_token(client_id=spotify_client_id,client_secret=spoitfy_client_secret)
```

```{r,echo=F}
access_token<-get_spotify_access_token(client_id=Sys.getenv("spotify_client_id"),client_secret=Sys.getenv("spotify_client_secret"))
```

```{r}
county_commute_plot_tunes <-function(artist,song,access_token){
  
  artists <- get_artists(artist,access_token=access_token)
  albums <- get_albums(artists$artist_uri[1],access_token=access_token)
  tracks<-get_album_tracks(albums,access_token=access_token)
  

  track= tracks[which(grepl(song,tracks$track_name)),][1,"track_uri"]
  
  audio_features <- get_track_audio_features(track,access_token=access_token)
  
  songLength=audio_features$duration_ms/1000/60
  
  all_county$commuteTune=all_county$MeanCommute/songLength
  
  ggplot(data=counties,mapping=aes(x=long,y=lat,group=group))+
    geom_polygon(data=all_county, aes(fill=commuteTune),color="grey")+labs(fill=paste("Song Plays Per \n Average Commute"))+scale_fill_distiller(palette="Spectral")+theme_void()+ggtitle(paste(artist,song,sep=" - "))+
    geom_path(data=state, aes(x=long,y=lat,group=group),color="black") ## add state boundaries
  
}

```

# Commuting 

What are these hot spots?

```{r,fig.width=12}
p=county_plot("MeanCommute") 
p
```

We can use `ggplotly` to use hover information to identify counties of interest.

```{r,eval=F}
county_plotly <-function(x){
  ## adapted from
  
  ##https://twitter.com/AidoBo/status/991338257391804416
  
  all_county$x<-all_county[,x] ## a different fix for this? something like aes_string?
  
  ggplot(data=counties,mapping=aes(x=long,y=lat,group=group))+
    geom_polygon(data=all_county, aes(fill=x,region=region,subregion=subregion),color="grey")+labs(fill=x)+scale_fill_distiller(palette="Spectral")+theme_void()+
    geom_path(data=state, aes(x=long,y=lat,group=group),color="black") ## add state boundaries
  
}

test=county_plotly("MeanCommute")
ggplotly(test,tooltip=c("region","subregion"))
```



**Must be the money?**

```{r,fig.width=12}
county_commute_plot_tunes("Nelly","Ride Wit Me",access_token)
```



**Where can commuting make you make more money?**

Caveat: I'm not *really* answering this question because we don't have the data at the individual level, but as an exploratory exercise...

```{r}
p <- ggplot(acs, aes(x = MeanCommute, y = IncomePerCap, text =paste(County,State,sep="-"))) +
  geom_point() +xlab("mean commute")+
  ylab("income per cap")+ggtitle("Where Does/Doesn't Commuting Pay Off?")
p ## static for GitHub
#ggplotly(p)
```

Can we get a rough idea of where it does and doesn't pay to commute on a map instead of relying on hovering?

```{r,fig.width=12}
## which counties have above average income and below average commute time per state (averages within a state)
averagesByState=group_by(acs,State)%>% summarize(avgMeanCommute=mean(MeanCommute),avgIncomePerCap=mean(IncomePerCap))


acsM=merge(acs,averagesByState,by.x="State",by.y="State",all.x=T)

acs$goodCommuteIncomeLevels=rep(0, nrow(acs))
acs$goodCommuteIncomeLevels[which(acsM$IncomePerCap>acsM$avgIncomePerCap & acsM$MeanCommute < acsM$avgMeanCommute)]=1
acs$goodCommuteIncomeLevels=as.factor(acs$goodCommuteIncomeLevels)

all_county<-inner_join(counties,acs %>% mutate(County=tolower(County),State=tolower(State)),by=c("subregion"="County","region"="State"))

## need a discrete version of the map
ggplot(data=counties,mapping=aes(x=long,y=lat,group=group))+
    geom_polygon(data=all_county, aes(fill=goodCommuteIncomeLevels),color="grey")+labs(fill="Good Commuter Given Income")+scale_fill_discrete()+theme_void()+
    geom_path(data=state, aes(x=long,y=lat,group=group),color="black") ## add state boundaries
  
```



# Work At Home

```{r,fig.width=12}
p=county_plot("WorkAtHome") 
p
```

Again we could use `ggplotly` to identify hot spots?

```{r,eval=F}
test=county_plotly("WorkAtHome")
ggplotly(test,tooltip=c("region","subregion"))
```

A lot could be going on here, so again I don't want to read too much into this plot. Since we don't have income per person we don't know if those working from home make more or less than those in other jobs within their county. However, there are some interesting patterns here that it would be interesting to look into with data at the individual level. 

```{r}

ggplot(acs,aes(x=WorkAtHome,y=acs$IncomePerCap))+geom_point() +xlab("percentage working from home")+
  ylab("income per cap")+ggtitle("Does Working From Home Pay?")

```

```{r,fig.width=12}
county_commute_plot_tunes("Fifth Harmony","Work from Home",access_token)
```


Note: My `county_commute_plot_tunes` is not robust to capitalization. There was some trial and error involved.