---
title: "The Bay Area's Five Star Sound"
date: 2019-10-29T10:20:51-07:00
---

I signed up to get Dan Kopf's "Golden Stats Warrior" newsletter, a newsletter that provides data driven insight into the Bay Area. You can sign up too [here](https://goldenstatswarrior.substack.com/).

The last edition looked at the greatest albums from the Bay Area. Of albums by Bay Area natives, which ones got five stars on [AllMusic](https://www.allmusic.com/)? Being a bit of a [music nerd](http://www.covermesongs.com/author/sarastoudt), I wanted to dig in further. Kopf was kind enough to share the data with me so that I could do some more exploring. 

I decided to use the Spotify API to get audio features for all of the songs on these albums. Then we can see if there is a quintessential Bay Area sound. 

```{r,warning=F, message=F}
library(dplyr)
library(spotifyr)
library(purrr)
library(lubridate)
library(ggplot2)
library(ggradar)
library(kableExtra)
library(scales)
```

```{r,eval=F}
spotify_client_id="" ## put yours here
spotify_client_secret="" ## put yours here
access_token <- get_spotify_access_token(client_id=spotify_client_id,client_secret=spotify_client_secret)
```

```{r}
setwd("~/Desktop/freelance/goldenStatWarrior")
data <- read.csv("BayAreaAlbums-datafromletter3.csv", stringsAsFactors = F)
greatest <- subset(data, Score == 5)
```

First we grab the audio features for any album by the Bay Area artists who scored a five on at least one album.

```{r,eval=F}
helper <- function(x) {
  get_artist_audio_features(x, include_groups = "album", authorization = access_token)
}



safe_spotify <- safely(helper) ## don't want it to crash if there is a bad request

allInfoGreatest <- unique(greatest$Artist.Name) %>% map(safe_spotify)
save(allInfoGreatest, file = "allInfoGreatest.RData")

testSly <- helper("family stone") ## Sly & the Family Stone gets weird, so do them seperately
```

Then we clean that up a bit.

```{r,eval=F}
load(file = "allInfoGreatest.RData")
clean <- lapply(allInfoGreatest, function(x) {
  x$result
})
clean2 <- do.call("rbind", clean)

clean2$album_release_date <- as.Date(clean2$album_release_date)
clean2$year <- year(clean2$album_release_date)
testSly$album_release_date <- as.Date(testSly$album_release_date)
testSly$year <- year(testSly$album_release_date)
clean3 <- rbind(clean2, testSly)

save(clean3, file = "clean3.RData")
```

Now we get things prepped for merging based on names. This could get ugly.

```{r}
setwd("~/Desktop/freelance/goldenStatWarrior")
load(file="clean3.RData")

greatest$Artist.Name=tolower(greatest$Artist.Name)
greatest$Album.Name=tolower(greatest$Album.Name)
clean3$album_name=tolower(clean3$album_name)
clean3$artist_name=tolower(clean3$artist_name)
```

Inevitably, it did. These were the ones where I had to do some manual fudging to get things right.

```{r}
greatest$Album.Name[which(greatest$Album.Name == "handel: messiah [2008 recording]")] <- "handel: messiah"
greatest$Artist.Name[which(greatest$Artist.Name == "dave brubeck")] <- "the dave brubeck quartet"

greatest$Album.Name[which(greatest$Album.Name == "lorraine at emmanuel")] <- "lorraine hunt lieberson at emmanuel"

greatest$Album.Name[which(greatest$Album.Name == "recital: lorraine hunt lieberson at ravinia")] <- "recital at ravinia"

greatest$Album.Name[which(greatest$Album.Name == "red house painters (roller-coaster)")] <- "red house painters i"

greatest$Album.Name[which(greatest$Album.Name == "john adams: harmonielehre; short ride in a fast machine")] <- "adams: harmonielehre - short ride in a fast machine"

greatest$Artist.Name[which(greatest$Artist.Name == "vince guaraldi")] <- "vince guaraldi trio"

greatest$Album.Name[which(greatest$Album.Name == "cast your fate to the wind: jazz impressions of black orpheus")] <- "cast your fate to the wind"

greatest$Album.Name[which(greatest$Album.Name == "a boy named charlie brown [original soundtrack]")] <- "a boy named charlie brown"

greatest2 <- rbind.data.frame(greatest, greatest[which(greatest$Album.Name == "cast your fate to the wind"), ])

greatest2$Album.Name[nrow(greatest2)] <- "jazz impressions of black orpheus" ## split into two on Spotify

```

Actually do the matching.

```{r}
helper <- function(x) {
  use <- greatest2[x, c("Album.Name", "Artist.Name", "Year")]

  test <- subset(clean3, artist_name == use$Artist.Name)
  toReturn <- test[agrep(use$Album.Name, test$album_name), ] ## sort of fuzzy match
  return(toReturn)
}


getData <- lapply(1:nrow(greatest2), helper)
test <- lapply(getData, nrow)
```

Which ones didn't work out?

```{r}
tt = greatest[which(unlist(test)==0),] 
kable(tt) %>% kable_styling()
```

I checked, and this wasn't just an issue of merging on slightly different names. I couldn't find the analogue in the albums pulled from Spotify based on the artist names. We only lost 16 out of 56, not bad!

```{r}
dataINeed=do.call("rbind",getData)
save(dataINeed,file = "fullData2018-10-28.RData")
```



Now we get to the fun stuff. What kind of features do we have? We want to get the average and variability across songs, per album.

```{r}
characteristics <- c("danceability", "energy", "loudness", "speechiness", "acousticness", "instrumentalness", "liveness", "valence")

charSum <- dataINeed[, c(characteristics, "album_name")] %>%
  group_by(album_name) %>%
  summarise(
    meanDanceability = mean(danceability), meanEnergy = mean(energy),
    meanLoudness = mean(loudness), meanSpeechiness = mean(speechiness),
    meanAcousticness = mean(acousticness), meanInstrumentalness = mean(instrumentalness), meanLiveness = mean(liveness), meanValence = mean(valence),
    varDanceability = var(danceability), varEnergy = var(energy),
    varLoudness = var(loudness), varSpeechiness = var(speechiness),
    varAcousticness = var(acousticness), varInstrumentalness = var(instrumentalness), varLiveness = var(liveness), varValence = var(valence)
  )


## add mean of everything in black
add <- cbind.data.frame(album_name = "avg", t(apply(dataINeed[, c(characteristics)], 2, mean)))


meanStuff <- charSum[, 1:9]
names(add) <- names(meanStuff)
meanStuff <- rbind.data.frame(meanStuff, add)
varStuff <- charSum[, c(1, 10:17)]

```

Now, I wanted to make some cool radar plots, inspired by music analysis like [this](https://www.freecodecamp.org/news/spotifys-this-is-playlists-the-ultimate-song-analysis-for-50-mainstream-artists-491882081819/). You can also find descriptions of the features there. 

Of course there is a `gg` version, but I had to tweak the internal functionality to get the plots to look like what I wanted them to. This gets a bit gnarly, so I'm compartmentalizing that code over [here](https://gist.github.com/sastoudt/fc4580ceb903add59df0b670a8451954).

```{r}
setwd("~/Desktop/freelance/goldenStatWarrior")
source("fix_ggradar.R")
```

I also snagged some code from the internal `ggradar` so that I could add a line representing the average across all albums in black. That stuff lives in the following helper function.

```{r}
helperPlot <- function(base) {
  grid.min <- 0 # , # 10,
  grid.mid <- 0.5 # , # 50,
  grid.max <- 1 # , # 100,
  centre.y <- grid.min - ((1 / 9) * (grid.max - grid.min))
  plot.data <- as.data.frame(toP[nrow(toP), ])
  names(plot.data)[1] <- "group"


  if (!is.factor(plot.data[, 1])) {
    plot.data[, 1] <- as.factor(as.character(plot.data[, 1]))
  }

  plot.data.offset <- plot.data
  plot.data.offset[, 2:ncol(plot.data)] <- plot.data[, 2:ncol(plot.data)] + abs(centre.y)
  # print(plot.data.offset)
  # (b) convert into radial coords
  group <- NULL
  group$path <- CalculateGroupPath(plot.data.offset)

  group.line.width <- 1.5
  group.point.size <- 6
  # ... + group (cluster) 'paths'
  base <- base + geom_path(
    data = group$path, aes(x = x, y = y, group = group),
    size = group.line.width, lty = 2, col = "black"
  )

  # ... + group points (cluster data)

  base <- base + geom_point(data = group$path, aes(x = x, y = y, group = group), size = group.point.size, alpha = 0.5, col = "black")

  return(base)
}
```

Acousticness for the win! However, there are quite a few albums that hit high instrumentalness (makes sense, lots of jazz and classical music), high energy, danceability (they aren't called Funky Divas for nothing), and loudness (Green Day, anyone?)

```{r,out.width = "80%", fig.height=10, fig.width=10}
toP <- meanStuff %>% mutate_at(vars(-album_name), rescale)

base <- ggradar2(toP[-nrow(toP), ])

helperPlot(base)
```

Which album is most like the average five-star Bay Area album? Dave Brubeck's "Time Out"	

This isn't particularly suprising has Brubeck has the third highest number of albums on the five-star list.

```{r}
test=as.matrix(dist(toP[,-1]))

#toP[which.min(test[nrow(toP),-ncol(test)]),]

dataINeed %>% group_by(album_id) %>% summarise(count=n(),artist=artist_name[1])  %>% group_by(artist) %>% summarise(count=n()) %>% arrange(desc(count))

```

The Bay Area likes variability in valence (positivity of sound).

```{r,out.width = "80%", fig.height=10, fig.width=10}
toP = varStuff %>% mutate_at(vars(-album_name), rescale) 

base = ggradar2(toP[-nrow(toP),])

helperPlot(base)
```

Which album has the variability most like the average five-star Bay Area album? The Grateful Dead's "American Beauty" (from SF)

```{r}
test=as.matrix(dist(toP[,-1]))

#toP[which.min(test[nrow(toP),-ncol(test)]),]

```

What about the [Bay Area pace](https://sastoudt.github.io/blog/songs-to-strut-to/)? Let's look at tempo.

```{r,warning=F,message=F}
ggplot(dataINeed, aes(x=tempo)) + geom_histogram()+xlab("beats per minute")
```

What song is most like the average tempo? Ironically, Brubeck's St. Louis Blues. Brubeck is a Concord original.

```{r}
mean(dataINeed$tempo)
#dataINeed[which.min(abs(dataINeed$tempo-mean(dataINeed$tempo))),]

blogdown::shortcode("youtube", "tq2YENV_Q9s")
```

But there is a peak in the tempo distribution that is slower than the mean. What song is most like this mode? Gotta love a good cover of The Beatles' "Yesterday" by Oakland's En Vogue.

```{r}
#which.max(table(round(dataINeed$tempo)))

# mode
#dataINeed[which.min(abs(dataINeed$tempo-80)),]

blogdown::shortcode("youtube", "k1PiJAeydLs")
```

What about those few faster paced jams? Here is an inspirational song from San Francisco's Moby Grape.

```{r}
#dataINeed[which.min(abs(dataINeed$tempo-200)),]

blogdown::shortcode("youtube", "9-RKXCvb5E")

```

What is an example of a song that is in the most prominent key? Oakland's Digital Underground gives us "Doowutchyalike."

```{r}
#which.max(table(dataINeed$key_mode))

#dataINeed[which(dataINeed$key_mode=="A major")[1],]

blogdown::shortcode("youtube", "5P4WZHlHsyk")
```

There is plenty more to dig into here. 

- What about all of the other star-levels? 
- Are the 5-star albums from another area noticeably different than the Bay Area ones?
- Do these Bay Area albums call out their roots in the song lyrics?

But those are for another day...

Thanks again to Dan Kopf (@dkopf) for sharing the data! Thoughts, comments, suggestions, etc. welcome --> @sastoudt.